Standartox: A tool to derive exposure endpoints from chemical test results

### 2019-11-11 article/00_article.tex
Updates from Overleaf

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage{natbib} % citations
\usepackage{csvsimple}
\usepackage{pgfplotstable}
\usepackage{rotating} % to rotate tables
%% tikz packages
\usepackage{tikz}
\usepackage{forest}
\usetikzlibrary{trees, arrows, shapes.geometric, positioning}
\usepackage{standalone}
%% text packages
\usepackage{siunitx} % fake ° symbol \ang{}
\usepackage[super]{nth} % for 1st, 2nd etc. \nth{}
\usepackage{hyperref}

\newcommand{\standartox}{STANDARTOX}
\newcommand{\epa}{EPA ECOTOX data base}
\newcommand{\rpackage}{http://139.14.20.252:3838/etox-base-shiny/}
\newcommand{\git}{https://github.com/andreasLD/etox-base}
\newcommand{\gitapp}{https://github.com/andreasLD/etox-base-shiny}
\newcommand{\ecfifty}{EC\textsubscript{50}}

@@

\begin{figure}
    \includestandalone[scale=0.5]{tikz/tree_source}
    \caption{Alternative processing pipeline}
    \label{fig:pipeline-tree}
\end{figure}

@@

\pagebreak
\section*{Supplement}
\subsection*{R-package list} %% TODO put in appropriate list
\label{list:r-packages}
%% TODO how to refernce this?
\input{article/tables/r_package_list.tex}
\pagebreak

\subsection*{Data base tables}
\input{article/tables/data_base_tables.tex}

\subsection*{Application parameters}
\input{article/tables/app_parameters.tex}

>>>>

\documentclass[english]{article}

%% ---------------------------- packages ------------------------------------ %%
\usepackage[T1]{fontenc}
\usepackage{natbib} % citations
\usepackage{csvsimple, longtable, booktabs}
\usepackage{pgfplotstable}
\usepackage{rotating} % to rotate tables
%% tikz packages
\usepackage{tikz}
\usepackage{forest}
\usetikzlibrary{trees, arrows, shapes.geometric, positioning}
\usepackage{standalone}
%% text packages
\usepackage{siunitx} % fake ° symbol \ang{}
\usepackage[super]{nth} % for 1st, 2nd etc. \nth{}
\usepackage{hyperref}
\usepackage{listings} % code highlighting
\usepackage{color}

%% ---------------------------- commands ------------------------------------ %%
\newcommand{\standartox}{STANDARTOX}
\newcommand{\epa}{EPA ECOTOX data base}
\newcommand{\rpackage}{http://139.14.20.252:3838/etox-base-shiny/}
\newcommand{\git}{https://github.com/andreasLD/etox-base}
\newcommand{\gitapp}{https://github.com/andreasLD/etox-base-shiny}
\newcommand{\ecfifty}{EC\textsubscript{50}}

%% ---------------------------- code ---------------------------------------- %%
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=R,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

@@

\begin{figure}
    % TODO make better graphic
    \includegraphics[width=1\linewidth]{article/figures/scripts_dendrogram.png}
    \caption{Processing pipeline}
    \label{fig:pipeline-tree}
\end{figure}

@@

\input{article/parts/06_supplement.tex}
\pagebreak

### 2019-11-11 article/parts/01_background.tex

The risk of adverse effects of chemicals on the environment is globally assessed with the help of ecotoxicological tests that are conducted on different organism groups, ideally representing a specific biocoenosis and associated biotopes. These tests are conducted by different research groups all over the world who publish the results independently. Up to now, only few initiatives, such as the United States Environmental Protection Agency ECOTOXicology knowledgebase (EPA ECOTOX) \citep{elonen_ecotoxicology_2018}, the ETOX (UBA ETOX) <cite UBA> or the Pesticide Property Data Base (PPDB) which aim to consolidate ecotoxicological test results, exist. However, the former two only collect test results and hence partly check for quality whereas the latter aims for quality, yet being confined to specific classes of chemicals, in this case pesticides. An overall holistic approach in processing the large number of ecotoxicological test results is lacking. Therefore, we suggest Standartox, a tool and data base that aims to overcome these paucities by steadily incorporating the ever-growing number of test results in an automated process workflow that ultimately leads to single aggregated test results for specific chemical-organism test combinations, representing the toxicity of a chemical. Automated approaches are increasingly important because a growing number of chemicals such as pharmaceuticals, pesticides and synthetic hormones, amongst others are in daily use all over the world with insufficient knowledge about possible adverse effects on the environment. Generally, chemicals are brought to the environment deliberately - as it is the case of pesticides, or arrive therein as a byproduct from other processes (e.g. atmospheric emissions or wastewater) \citep{schwarzenbach_challenge_2006}. In turn, ecosystems provide essential services to human societies such as drinking and irrigation water, food and climate regulation. These services are products of different ecosystem functions that crucially depend on the integrity of the populations and communities that drive these ecosystem functions. However, besides habitat degradation, climate change and nutrient enrichment, pollution with man-made chemicals threatens these populations and communities in various ways which are currently not fully understood \citep{steffen_anthropocene_2007}. Pollution with man-made chemicals was indeed identified as one of three major environmental problems for which research gaps hamper the derivation of planetary boundaries, i.e. thresholds beyond which irreversible state shifts may occur \citep{steffen_anthropocene_2007}. Bernhardt et al. \citet{bernhardt_synthetic_2017} argue further that the knowledge gap how chemicals effect populations and communities and hence ecosystem functions and ecosystem services, would also impede society's ability to accomplish the Sustainable Development Goals of the United Nations. In Europe alone some 100,000 chemicals are estimated to be in current use, whereof 30,000 are produced in quantities larger than one ton per year \citep{breithaupt_costs_2006}. Data on toxicity tests for a lot of chemicals, especially newly emerging ones remains scarce and if existent, the information is represented sparsely and in inconsistent formats \citep{gessner_fostering_2016}. Therefore, tools such as Standartox are a needed step adding a layer of abstraction satisfying modern environmental risk assessment demands. Standartox aggregates ecotoxicological test results in a standardized and reproducible way, providing the basis for the derivation of risk indicators such as Species Sensitivity Distributions (SSDs) \citep{posthuma_species_2002} and Toxic Units (TUs), which represent two prominent concepts to assess effects on organisms in ecotoxicology \citep{kefford_definition_2011, schafer_effects_2011}. Besides aggregating ecotoxicological test results, Standartox provides a concise overview of the status of tested chemicals, allowing the identification of potential research gaps. Furthermore Standartox comes as an R package largely reducing search time and adding convenience structures for researchers \citep{rcoreteam_language_2017}.

@@

include \citep{yamamuro_neonicotinoids_2019} as a new prominent examples for effects on the ecosystem

>>>>

The risk of adverse effects of chemicals on the environment is globally assessed with the help of ecotoxicological tests that are conducted on different organism groups, ideally representing a specific biocoenosis and associated biotopes. These tests are conducted by different research groups all over the world who publish the results independently. Up to now, only few initiatives, such as the United States Environmental Protection Agency ECOTOXicology knowledgebase (EPA ECOTOX) \citep{elonen_ecotoxicology_2018}, the German Umweltbundesamt's Information System Ecotoxicology and Environmental Quality Targets (ETOX) \citep{umweltbundesamt_etox_2019} or the Pesticide Property Data Base (PPDB) aiming to consolidate ecotoxicological test results, exist. The former two collect test results and hence only partly check for quality whereas the latter aims for quality, yet being confined to specific classes of chemicals, in this case pesticides. An overall holistic approach in processing the large number of ecotoxicological test results is lacking. Therefore, we suggest Standartox, a tool and data base that aims to overcome these paucities by steadily incorporating the ever-growing number of test results in an automated process workflow that ultimately leads to single aggregated test results for specific chemical-organism test combinations, representing the toxicity of a chemical. Automated approaches are increasingly important because a growing number of chemicals such as pharmaceuticals, pesticides and synthetic hormones, amongst others are in daily use all over the world with insufficient knowledge about possible adverse effects on the environment. Generally, chemicals are brought to the environment deliberately - as it is the case of pesticides, or arrive therein as a byproduct from other processes (e.g. atmospheric emissions or wastewater) \citep{schwarzenbach_challenge_2006}. In turn, ecosystems provide essential services to human societies such as drinking and irrigation water, food and climate regulation. These services are products of different ecosystem functions that crucially depend on the integrity of the populations and communities that drive these ecosystem functions. However, besides habitat degradation, climate change and nutrient enrichment, pollution with man-made chemicals threatens these populations and communities in various ways which are currently not fully understood \citep{steffen_anthropocene_2007}. Pollution with man-made chemicals was indeed identified as one of three major environmental problems for which research gaps hamper the derivation of planetary boundaries, i.e. thresholds beyond which irreversible state shifts may occur \citep{steffen_anthropocene_2007}. Bernhardt et al. \citet{bernhardt_synthetic_2017} argue further that the knowledge gap how chemicals effect populations and communities and hence ecosystem functions and ecosystem services, would also impede society's ability to accomplish the Sustainable Development Goals of the United Nations. In Europe alone some 100,000 chemicals are estimated to be in current use, whereof 30,000 are produced in quantities larger than one ton per year \citep{breithaupt_costs_2006}. Data on toxicity tests for a lot of chemicals, especially newly emerging ones remains scarce and if existent, the information is represented sparsely and in inconsistent formats \citep{gessner_fostering_2016}. Therefore, tools such as Standartox are a required step adding a layer of abstraction satisfying the needs of modern environmental risk assessment demands. Standartox aggregates ecotoxicological test results in a standardized and reproducible way, providing the basis for the derivation of risk indicators such as Species Sensitivity Distributions (SSD) \citep{posthuma_species_2002} and Toxic Units (TU), which represent two prominent concepts to assess effects on organisms in ecotoxicology \citep{kefford_definition_2011, schafer_effects_2011}. Besides aggregating ecotoxicological test results, Standartox provides a concise overview of the status of tested chemicals, allowing the identification of potential research gaps. Standartox comes with two front-ends, an R \citep{rcoreteam_language_2017} package and a web tool, adding convenience structures and largely reducing work time for users.

@@

include \citep{yamamuro_neonicotinoids_2019} and \citep{seibold_arthropod_2019} as a new prominent examples for effects on the ecosystem

### 2019-11-11 article/parts/04_technical_validation.tex

%%This section presents any experiments or analyses that are needed
%%to support the technical quality of the dataset. This section may
%%be supported by up figures and tables, as needed. This is a required
%%section; authors must present information justifying the reliability
%%of their data.

\section{Technical validation}

\subsection{Comparison to other data bases}

In order to asses the validity of the Standartox approach to aggregate test results we compared Standartox values with effect values for the same chemicals and the same test duration from the PPDB (n=3601) \ref{fig:standartox_ppdb_diff}. This showed that 91.3\% of the Standartox values diverge by less than a factor of 10 from the PPDB values. Considering only Standartox values for which at least five of ten tests were aggregated the value increases to 91.7\% and 93.4\% respectively. Admittedly a few Standartox values diverge by a large margin.

@@

\subsection{Unit Harmonisation}

In order to assure appropriate conversion and harmonisation of units we took a random sample of the 50 most occurring units and converted them manually and subsequently compared them to the conversions done by Standartox.

We did this for concentration units and duration units

in R/chck\_unit\_conversion.R

>>>>

%%This section presents any experiments or analyses that are needed
%%to support the technical quality of the dataset. This section may
%%be supported by up figures and tables, as needed. This is a required
%%section; authors must present information justifying the reliability
%%of their data.
\section{Technical validation}

\subsection{Unit Harmonisation}
In order to guarantee appropriate conversion and harmonisation of units we take a random sample of the 50 most occurring concentration units (Table \ref{tab:chck-concentration}) as well as six of the most occurring duration units (Table \ref{tab:chck-duration}) and convert them manually to subsequently compare them to the conversions done by Standartox. This is check is done whenever Standartox is rebuilt.

\subsection{Comparison to other data bases}
In order to asses the validity of the Standartox approach to aggregate test results we compared Standartox values with effect values for the same chemicals and the same test duration from the PPDB (n=3601) \ref{fig:standartox_ppdb_diff}. This showed that 91.3\% of the Standartox values diverge by less than a factor of 10 from the PPDB values. Considering only Standartox values for which at least five of ten tests were aggregated the value increases to 91.7\% and 93.4\% respectively. Admittedly a few Standartox values diverge by a large margin.

@@

### 2019-11-11 article/parts/05_usage_notes.tex

\section*{Usage Notes}
%%Brief instructions that may help other researchers reuse these dataset.
%%This is an optional section, but strongly encouraged when helpful
%%to readers. This may include discussion of software packages that
%%are suitable for analyzing the assay data files, suggested downstream
%%processing steps (e.g. normalization, etc.), or tips for integrating
%%or comparing this with other datasets. If needed, authors are encouraged
%%to provide code, programs, or data processing workflows when they may help 
%%others analyse the data. We encourage authors to archive related code in 
%%a DOI-issuing archive when possible, but code may also be supplied as 
%%supplementary information files. 
%%
%%For studies involving privacy or safety controls on public access
%%to the data, this section should describe in detail these controls,
%%including how authors can apply to access the data, and what criteria
%%will be used to determine who may access the data, and any limitations
%%on data use.

\subsection*{Web application}

Users can define several input parameters \ref{tab:inputs} in the application's graphical user interface (GUI), handled in the ui.R script in order to adjust the aggregation according specific requirements. In doing so users execute R functions (fun\_filter.R, fun\_aggregation.R, fun\_filagg\_plot\_ly.R) loaded in the server.R script, which produce a 'filtered' and an 'aggregated' data set as well as an interactive overview plot. Only the 'aggregated' data set is presented in the GUI. However, both data sets can be downloaded via a download button.

The \etoxbase{} is designed to support CRA of ecotoxicologists. With an increased amount of available ecotoxicological test data, it becomes fundamental to provide and distribute such information in reasonable formats, meaning accessible for humans as well as machines. Currently CRA often relies on test results of a few, well tested standard organism, such as \textit{Daphnia magna}, \textit{Pimephales promelas}, \textit{Selenastrum capricornutum}, although researchers have conducted experiments on a much greater variety of organisms (e.g. the 926,108 test results collected in the EPA ECOTOX data base). In order to locate ecotoxicological test data, researchers can rely on individual test results reported in publications <CITE>, on compiled and published data sets \citep{malaj_organic_2014} or on data bases such as the Pesticide Property Data Base (PPDB) \citep{lewis_international_2016}. Each data source has its limitations though: Collecting toxicity information from individual publications is laborious and using already compiled data from publications or data bases is often limited to a specific groups of chemicals, such as pesticides, biozides etc. The \etoxbase{}, making use of the EPA ECOTOX data base, the largest publicly available collection of ecotoxicological test results, aims to facilitate the data retrieval process by providing adequate aggregates (\ecfifty{}, NOEC, LOEC) of the tests results. These are then used for the calculation of SSDs, TUs or other risk indicators. SSDs try to extrapolate effects of chemicals towards organisms from species to a community level by fitting a statistical distribution to toxicity test results on organisms of different trophic levels in order to derive concentrations at which a specific fraction of organisms are effected (often at 5\% - HC\textsubscript{5} \citep{posthuma_species_2002}. Figure \ref{fig:ssd-isoproturon} shows an example of a SSD and the derived HC\textsubscript{5} value (red dassed line) created for the herbicide Isoproturon. TUs are another indicator of effects of chemicals on organisms. They are calculated by dividing a measured concentration (c) by an \ecfifty{} value (cf. Equation \ref{eq:tu}).

@@

\begin{figure}[h!]
    \dentering
    \includegraphics[width=1\linewidth]{article/figures/ssd2_boot.png}
    \caption{SSD plot showing the susceptibility of organisms towards the herbicide Isoproturon. The red line represents the fit and the red dashed line marks the HC\textsubscript{5} value, which is a common measure in CRA. The legend denotes common organism groups used in ecotoxicology.}
    \label{fig:ssd-isoproturon}
\end{figure}

\subsection*{Filter and Aggregate}

In the application's GUI, the user can choose among different parameters to filter and aggregate the test data accordingly \ref{tab:inputs}. Percentage values next to inputs show the proportion of the respective option (e.g. \textit{Active Ingredient - 43\%} indicates that 43\% of the data are Active Ingredients). The user can filter for compound-specific filters, including the CAS number, the concentration type (e.g. Active Ingredient, Formulation), the chemical class (e.g. Insecticides, Metals) and taxon-specific filters, including common taxonomic groups (e.g. Daphniidae, Algae), the organism habitat (e.g. freshwater) and the continent (e.g. Europe, Asia). Regarding the test parameters the user can choose test durations (in hours), Effect groups (e.g. Mortaliy, Population, Growth) and Endpoints (\ecfifty, EC\textsubscript{10}, LOEC or NOEC). Besides that, two cleaning steps can also be chosen. Firstly the application allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective chemical at \ang{20} C. Secondly the user can choose to exclude outliers (cf. Table \ref{tab:scripts-app}). The outliers are selected as values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the inter-quartile range. As aggregates the user can choose the minimum, the maximum, the median, the geometric mean or the arithmetic mean. All the steps described in this section are selectable in the GUI and are executed at each click by the application. The drop down menu `Download data` allows to download a filtered data set as well as an aggregated data set.
The \etoxbase{} application can be accessed here: \app{}.

\begin{table}[h!]
\begin{tabular}{|l|l|l|l|}
\hline
Menu   & Group       & Subgroups              & Options \\ \hline
Filter      & Compound    & CAS                    & input CAS \\ \hline
Filter      & Compound    & Concentration type     & Active Ingredient; Formulation; Total \\ \hline
Filter      & Compound    & Chemical class         & Insecticides; Herbicides; Metals \\ \hline
Filter      & Taxon       & Taxon                  & Choose a taxon \\ \hline
Filter      & Taxon       & Organism habitat       & freshwater; terrestrial; marine; brackish \\ \hline
Filter      & Taxon       & Continent              & North America; Asia; Europe \\ \hline
Filter      & Test        & Duration               & enter duration in hours \\ \hline
Filter      & Test        & Effect groups          & MOR - Mortality; POP - Population \\ \hline
Filter      & Test        & Endpoins               & EC50, LOEC, LOEL \\ \hline
Filter      & Checks      & Water solubility check & yes/no \\ \hline
Filter      & Checks      & Remove outliers        & yes/no \\ \hline
Aggregation & Aggregation & Aggregate              & Minimum; Maximum; Median \\ \hline
\end{tabular}
\caption{\etoxbase GUI paramters to choose.}
\label{tab:inputs}
\end{table}

\subsection*{Limitations}

Generally it has to be stated that the returned values are subject to change, as new tests will be constantly incorporated into the \etoxbase{}. However, we expect values of well tested chemical-organism combinations to be relatively stable. The returned values are certainly influenced by individual tests characteristics, such as pH, temperature or conductivity to just name a few, which influences the effect relationship between chemicals and organisms \citep{rosenkrantz_influence_2013,li_temperature_2011}. Unfortunately the EPA ECOTOX data base has only rare records on test parameters, which prevents incorporating such parameters in an aggregation <proportions of available data: Temperature:77\%, pH:56\%, hardness:27\%, dissolved oxygen:18\%, Alkalinity:15\%, salinity:9\%, others:<=5\%>. However, using the median or the geometric mean as an aggregate allows for an adequate estimation of the central tendency of the toxicity of a chemical towards an organism group. The geometric mean is recommended, as it is in comparison to the arithmetic mean considerably less influenced by outliers and is suitable for skewed data. Also, the geometric mean is preferable in contrast to the median, since the median completely ignores the influence of large or small values, making it unreliable for small data sets <CITATION GEOMETRIC MEAN>.

Other initiatives, with different, though partly overlapping aims accumulating existing ecotoxicological test data are published online. The PPDB provides validated ecotoxicological test results for pesticides, narrowing chemicals to this group \citep{lewis_international_2016}. The Network of reference laboratories, research centres and related organisations for monitoring of emerging environmental substances (NORMAN) focuses on assembling river basin specific pollutants\citep{von_der_ohe_new_2011}. The EnviroTox data base \href{https://envirotoxdatabase.org/} makes also use of the \epa{} data in order to derive Predicted no-effect concentrations (PNEC) \citep{health_and_environmental_sciences_institute_hesi_envirotox_2019}. In comparison to the \etoxbase, these examples partly rely on the same data, yet not aiming to derive \ecfifty, LOEC and NOEC from it automatically.

\pagebreak

>>>>

\section*{Usage Notes}
%%Brief instructions that may help other researchers reuse these dataset.
%%This is an optional section, but strongly encouraged when helpful
%%to readers. This may include discussion of software packages that
%%are suitable for analyzing the assay data files, suggested downstream
%%processing steps (e.g. normalization, etc.), or tips for integrating
%%or comparing this with other datasets. If needed, authors are encouraged
%%to provide code, programs, or data processing workflows when they may help 
%%others analyse the data. We encourage authors to archive related code in 
%%a DOI-issuing archive when possible, but code may also be supplied as 
%%supplementary information files. 
%%
%%For studies involving privacy or safety controls on public access
%%to the data, this section should describe in detail these controls,
%%including how authors can apply to access the data, and what criteria
%%will be used to determine who may access the data, and any limitations
%%on data use.
Standartox can either be accessed through the APP or the API together with the R-package standartox. Both access pathways allow for the same filters and aggregation methods (Table \ref{tab:app-parameters} to be applied on the data. In the former the user can download the resulting data sets as a comma-separated values file whereas in the latter the user can directly load them in R. The user can choose among different parameters to filter and aggregate the test data accordingly \ref{tab:app-parameters}. The user can filter for compound-specific filters, including the CAS number, the concentration type (e.g. Active Ingredient, Formulation), the chemical class (e.g. Insecticides, Metals) and taxon-specific filters, including common taxonomic groups (e.g. Daphniidae, Algae), the habitat (e.g. freshwater) and the regional occurrence (e.g. Europe, Asia) of the organisms. Furthermore, the user can refine the results to specific test durations (in hours), effect groups (e.g. Mortaliy, Population, Growth) and endpoints (EC\textsubscript{50}, EC\textsubscript{10}, LOEC or NOEC).  As aggregates the user can choose the minimum, the maximum, the median, the geometric mean or the arithmetic mean. All the steps described in this section are selectable in the GUI and are executed at each click by the application. The drop down menu `Download data` allows to download a filtered data set as well as an aggregated data set.

Percentage values next to inputs show the proportion of the respective option (e.g. \textit{Active Ingredient - 43\%} indicates that 43\% of the data are Active Ingredients).

\subsection*{R package}
The R-package standartox consists of the two functions stx\_catalog() and stx\_query(). The former allows the retrieval of a catalog of possible parameters that can be used as an input for the latter. stx\_query() fetches the toxicity values from the data base and returns them as a R list object (compare Code \ref{listing:standartox-example}).

\begin{lstlisting}[
    caption = {standartox R-package code} example,
    label = {listing:standartox-example}]
# install
install.packages('remotes')
remotes::install_github('andschar/standartox') # package not yet on CRAN
# retrieve catalog    
require(standartox)
catal = stx_catalog()
# retrieve data
l = stx_query(cas = '1071-83-6',
                endpoint = 'XX50',
                taxa = 'Oncorhynchus',
                duration = c(24, 120))
\end{lstlisting}

Standartox is designed to support CRA. With an increased amount of available ecotoxicological test data, it becomes fundamental to provide and distribute such information in reasonable formats, meaning accessible for humans as well as machines. Currently CRA often relies on test results of a few, well tested standard organism, such as \textit{Daphnia magna}, \textit{Pimephales promelas}, \textit{Selenastrum capricornutum}, although researchers have conducted experiments on a much greater variety of organisms (e.g. the 926,108 test results collected in the EPA ECOTOX data base). In order to locate ecotoxicological test data, researchers can rely on individual test results reported in publications <CITE>, on compiled and published data sets \citep{malaj_organic_2014} or on data bases such as the Pesticide Property Data Base (PPDB) \citep{lewis_international_2016}. Each data source has its limitations though: Collecting toxicity information from individual publications is laborious and using already compiled data from publications or data bases is often limited to a specific groups of chemicals, such as pesticides, biozides etc. Standartox, making use of the EPA ECOTOX, the largest publicly available collection of ecotoxicological test results, aims to facilitate the data retrieval process by providing adequate aggregates (EC\textsubscript{50}, NOEC, LOEC) of the tests results. These are then used for the calculation of SSDs, TUs or other risk indicators. SSDs try to extrapolate effects of chemicals towards organisms from species to a community level by fitting a statistical distribution to toxicity test results on organisms of different trophic levels in order to derive concentrations at which a specific fraction of organisms are effected (often at 5\% - HC\textsubscript{5} \citep{posthuma_species_2002}. Figure \ref{fig:ssd-isoproturon} shows an example of a SSD and the derived HC\textsubscript{5} value (red dassed line) created for the herbicide Isoproturon. TUs are another indicator of effects of chemicals on organisms. They are calculated by dividing a measured concentration (c) by an EC\textsubscript{50} value (cf. Equation \ref{eq:tu}).

@@

\begin{figure}[h!]
    \includegraphics[width=1\linewidth]{article/figures/ssd2_boot.png}
    \caption{SSD plot showing the susceptibility of organisms towards the herbicide Isoproturon. The red line represents the fit and the red dashed line marks the HC\textsubscript{5} value, which is a common measure in CRA. The legend denotes common organism groups used in ecotoxicology.}
    \label{fig:ssd-isoproturon}
\end{figure}

\subsection*{Limitations}
Since the EPA ECOTOX data at hand merely represents a sample of the reality of the toxicity of chemicals on organisms, it has to be stated that the returned Standartox values are subject to change. New studies will be published and constantly incorporated into Standartox, improving the toxicity estimates. However, we expect values of well tested chemical-organism combinations to be relatively stable. A future analysis on the smallest number of tests needed to return trustworthy results, could shed more light on this aspect. Furthermore, toxicity test results are certainly influenced by individual tests parameters, such as pH, temperature or conductivity amongst others \citep{rosenkrantz_influence_2013, li_temperature_2011}. Such information is not included in the aggregation performed by Standartox, since the EPA ECOTOX only provides sparse records on those. Most frequently provided test parameters (and their percentage) are medium temperature (77\%), pH:56\%, hardness:27\%, dissolved oxygen:18\%, Alkalinity:15\% and salinity:9\%. Others are provided for less than 5\% of the tests. A test-mining approach, iterating through the individual test studies could potentially increase this number. However, we argue that using the geometric mean as an aggregate allows for an adequate estimation of the toxicity of a chemical towards an organism group. We prefer the geometric mean in comparison to the arithmetic mean, since it is considerably less influenced by outliers and is suitable for skewed data. Also, the geometric mean is preferable over the median, since the median completely ignores the influence of large or small values, making it unreliable for small data sets <CITATION GEOMETRIC MEAN>.

\subsection*{Other data bases}
Other initiatives tidying the large number of ecotoxicological test results were also developed recently. They partly aim for overlapping goals, yet have limitations or objectives that differentiate them from Standartox. Since 2006, the PPDB provides validated ecotoxicological information on pesticides \citep{lewis_international_2016}.
% TODO lookup NORMAN citation
The Network of reference laboratories, research centres and related organisations for monitoring of emerging environmental substances (NORMAN) focuses on assembling river basin specific pollutants \citep{von_der_ohe_new_2011}. 
Connors et al. \citet{healthandenvironmentalsciencesinstitutehesi_envirotox_2019, connors_creation_2019} published the EnviroTox data base \href{https://envirotoxdatabase.org/} which also uses, amongst others the EPA ECOTOX data as an input. In comparison to Standartox, ENviroTox focuses only on aquatic organisms and uses an qualitative algorithm to exclude unreliable test results. They filter for example the data to fish, amphibian invertebrate and algae taxa, to test durations of above 24 hours and aquatic taxa only and add additional information on toxicity endpoints, such as chronic or acute tests as well as mode of action assignments. In contrast, Standartox doesn't refine to specific test durations, but leaves it up to the user to decide on such parameters. The EnvirTox data base also allows for an aggregation of test results to derive single toxicity values for individual taxa whereas Standartox performs this aggregation for individual chemicals. This has to be done in a cumbersome way by down- and uploading the data manually. Comptox, is a web tool published by the EPA which, similar to Standartox allows for filtering test results and the retrieval of additional chemical information. However, it lacks the possibility to aggregate toxicity test results \citep{usenvironmentalprotectionagency_comptox_2019}. Petschick et al. (in submission) modeled risk threshold level equivalents for aquatic organisms by using ecotoxicological effect data from the EPA ECOTOX. However, none of the above mentioned approaches aim for a standardized aggregation method of toxicity endpoints for individual chemicals. Besides, they also lack the possibility for automated and scriptable user requests for high level programming languages, such as R or Python. Along with newly created ecotoxicological data bases here, methods of how to efficiently store ecotoxicological data are also proposed as can be seen in the MAGIC Knowledge Base \citep{bub_graphing_2019}. This recent increase of efforts to compile ecotoxicological test data bases clearly emphasizes the need for holistic and automated analyses of large-scale ecotoxicological data. Standartox is a tool that puts its focus on the aggregation of toxicity data and by this adds a piece to the puzzle of modern ecotoxicological analyses.

\subsection*{Identify chemicals to be tested}
There is a clear need in ecotoxicology for harmonised approaches to collect ecotoxicological data as other data base compilation attempts recently published show. Hitchock et al. \citet{hitchcock_improving_2018} argue for an increased awareness concerning newly published ecotoxicological test results to facilitate meta analyses across studies. Important statistical parameters, such as mean estimates, variances and sample sizes. Likewise, state variables, including medium temperature, pH, mineral contents or organism age and sex should be reported (if not in the text, in an online repository). Standartox doesn't include such information mainly because the EPA ECOTOX reports only a few of these variables. A future research aim could be to use machine learning techniques to retrieve this information from the referenced articles.
Above all, recent studies showed that, in defiance of many regulation efforts \citep{schafer_future_2019} chemicals can still cause strong adverse effects on organisms and in turn the environment. Neonicotinoids, a group of insecticides was for example related to declines in insectivorous birds \citep{hallmann_declines_2014} and fish yields due to cascading food web effects \citep{yamamuro_neonicotinoids_2019}. Standartox provides means to synthesise ecotoxicological test data, thereby identify research gaps. Furthermore, standardized approaches could potentially reduce unnecessary animal testing by synthesizing existing ecotoxicological test data.

\begin{figure}
    \includegraphics[width=1\linewidth]{article/figures/heatmap_tests_n.png}
    \caption{30 most tested chemicals in EPA ECOTOX.}
    \label{fig:standartox_ppdb_diff}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TODO %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
1) \citep{staveley_challenge_2016} describe what is necessary for future ecotoxicology.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% OLD %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\iffalse
Besides that, two cleaning steps can also be chosen. Firstly the application allows to exclude test results exhibiting concentrations that are higher than the actual water solubility of the respective chemical at \ang{20} C. Secondly the user can choose to exclude outliers (cf. Table \ref{tab:scripts-app}). The outliers are selected as values that exceed lower (0.25) and the upper (0.75) quartile by 1.5 times the inter-quartile range.
\fi
